---
title: "3: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1. Discuss and navigate different data types in R
2. Create, manipulate, and explore datasets
3. Call packages in R

## Data Types in R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.

* **Matrices** 2 dimensional structure that contains elements of the same type.

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.

* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ---- #this pattern, # followed by 4 - allows you to expand and collapse this section
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector. c means concatenate
vector1
vector2 <- c("one","two","three") # character vector. We needed quotations b/c they're strings, or r will thing it's an operation.
vector2
vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector. Since TRUE and FALSE are capital, r knows its a logical statement, no need for quotations
vector3

vector1[3] # R chooses the 3rd item in vector1. [] means matrix subsetting, call up the ordinal number inside the brackets

# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # matrices need rows and columns. matrix is a function, r knows what it needs to do. 
matrix1
matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #byrow means that we start counting across rows, rather than going down a column.
matrix2
matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #dimnames means dimension names, so that it gives each column a header
matrix3
matrix1[4, ] #Gives every value in row 4
matrix1[ , 3] #Gives every value in column 3
matrix1[c(12, 14)] #Gives the values of numbers in position 12 and 14. Specific objects in the matrix
matrix1[c(12:14)] # Gives every value between position 12 and 14
matrix1[2:4, 1:3] #Gives rows 2-4, columns 1-3

cells <- c(1, 26, 24, 68) #building a matrix from scratch, starting with a vector
rnames <- c("R1", "R2")
cnames <- c("C1", "C2") 
matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # making a matrix from those pieces we just created
matrix4

# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1 #semicolon means you can run with just one keystroke. List is a collection of the matrices, vectors, etc that we wrote
list1[[2]]

# Data Frames ----
d <- c(1, 2, 3, 4) # What type of vector? Num
e <- c("red", "white", "red", NA) # What type of vector? Chr
f <- c(TRUE, TRUE, TRUE, FALSE) # What type of vector? Logi
dataframe1 <- data.frame(d,e,f) #
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # 

dataframe1[1:2,] # rows 1-2. R will always skip the first/header row.
dataframe1[c("ID","Passed")] # Only showing columns with ID and Passed
dataframe1$ID #$ calls up a specific column

```
Question: How do the different types of data appear in the Environment tab?

> Answer: dataframes, lists, matrices appear as data, vectors appear as values

Question: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

> Answer: It gives me a choice of the three variables, ID, color, and Passed.

### Coding challenge

Find a ten-day forecast of temperatures (Fahrenheit) for Durham, North Carolina. Create two vectors, one representing the high temperature on each of the ten days and one representing the low.

```{r}
hiF<-c(62,45,50,50,40,40,45,52,54,58)
loF<-c(31,29,43,24,23,23,26,30,42,41)
```

Now, create two additional vectors that include the ten-day forecast for the high and low temperatures in Celsius.

```{r}
hiC<-(hiF-32)*(5/9)
loC<-(loF-32)*(5/9)
```

Combine your four vectors into a data frame and add informative column names.

```{r}
dataframeTemp <- data.frame(hiF,hiC,loF,loC) 
names(dataframeTemp) <- c("HiF","HiC","LowF","LowC"); View(dataframeTemp) 
```

Use the common functions `summary` and `sd` to obtain basic data summaries of the ten-day forecast. How would you call these functions differently for the entire data frame vs. a single column? Attempt to demonstrate both options below.

```{r}
summary(dataframeTemp)
summary(dataframeTemp$HiF)
sd(dataframeTemp)#should only work for specific lists/columns
sd(dataframeTemp$HiC)
```


## Packages

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your libaray already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a package is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.


```{r}
# We will use the packages dplyr and ggplot2 regularly. 
#install.packages("dplyr") 
#install.packages("ggplot2")
# comment out install commands, use only when needed and re-comment

library(dplyr)
library(ggplot2)

# Some packages are umbrellas under which other packages are loaded
#install.packages("tidyverse")
library(tidyverse)
```

Question: What happens in the console when you load a package?

> Answer: It attaches other packages, and lists some conflicts


## Tips and Tricks

* Sequential section headers can be created by using at least four -, =, and # characters.

* The command `require(packagename)` will also load a package, but it will not give any error or warning messages if there is an issue.

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 

* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out. 

* Update your packages regularly! 